Efficient pedigree recording for fast population genetics simulation

### 2017-12-12 forwards_paper.tex
Update intro and tree sequence section.

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent,
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, Allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\approx 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

@@

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we describe how algorithmic tools and data structures developed for the
coalescent simulator \msprime{}
can be used to efficiently record, and later process, this history.
In so doing we record the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time including the genetic outcomes of each ancestral
meiosis.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree sequence}~\citep{kelleher2016efficient}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the \emph{sampled} individuals,
which results in considerable savings.
Another way of representing this same information
is known as the \emph{ancestral recombination graph} --
the {ARG} \citep{griffiths1997ancestral},
where nodes are common ancestor or recombination events --
which has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
We use tree sequences for the substantial algorithmic advantages exploited in~\citep{kelleher2016efficient}.

@@

The simulations above recorded information about each new individual 
in collection of tables that together encode a \emph{tree sequence}.
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
It is efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants are recorded as \emph{sites} and \emph{mutations}.
The format for these four tables is given in more detail in the Methods;
here we explain conceptually,
using the example of Figure~\ref{fig:example_tree_sequence}. This formulation
is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a much larger class of tree topology.

>>>>

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent,
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, Allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

@@

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we describe how algorithmic tools and data structures developed for the
coalescent simulator \msprime{}
can be used to efficiently record, and later process, this history.
In so doing we record the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time including the genetic outcomes of each ancestral
meiosis.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree sequence}~\citep{kelleher2016efficient}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the extant population. We show that we need only $O(N \log N)$
space to store the resulting tree sequence for a single crossover per
generation Wright-Fisher model.

@@

The simulations above recorded information about each new individual
in collection of tables that together encode a \emph{tree sequence}.
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
A closely related structure is known as the \emph{ancestral recombination graph} --
the {ARG} \citep{griffiths1991two,griffiths1997ancestral},
where nodes are common ancestor or recombination events --
which has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, the properties of the ARG as a computational structure have not
been studied and, despite several efforts to standardise a common
format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper.
ARGs are rarely used
in practise. In contrast, the algorithmic properties of tree sequence
algorithms have been explored in detail~\citep{kelleher2016efficient},
contributing substantially to the efficiency of the \msprime\ coalescent simulator.

Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants are recorded as \emph{sites} and \emph{mutations}.
The format for these four tables is given in more detail in the Methods;
here we explain conceptually,
using the example of Figure~\ref{fig:example_tree_sequence}. This formulation
is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a much larger class of tree topology.
